Deep mutational scanning of hemagglutinin helps predict evolutionary fates of human H3N2 influenza variants

### 2018-06-23 paper/reviewer_response_PNAS.tex

Of note, position 1 for the H1 HA in the supplementary figure is not even shown.

\response{
The reviewer is correct. 
Figure~S3 shows data for a H1 HA as measured previously by Doud and Bloom (2016), and these prior data are simply presented here in H3 numbering for comparative purposes (the original Doud and Bloom paper used a different HA numbering scheme that makes site-to-site comparisons with the current study difficult). 
The Doud and Bloom study used mutant libraries in which all codons in HA \emph{except} for the one encoding the N-terminal methionine were mutagenized.
Therefore, that study did not estimate the effects of mutations to the first codon, which is why it is not shown in Figure~S3.
We have added a sentence to the legend for this supplementary figure explaining this fact.
}

2) While other important features of the HA like several cysteines seem to be conserved, there is no pressure on other elements which are considered important for viral fitness including the NGT glycosylation site in the beginning of HA1 which is conserved in all group 2 HAs or the adjacent NAT glycosylation site which is conserved in H3, H7, H10 and H15. On the other hand, head glycosylation sites, which are highly variable in nature, seem to be conserved. Again, viruses without these glycosylation sites should be rescued and evaluated for fitness relative to wild type.

\response{This is a valid point, and we have now also individually evaluated the reverse-genetics titers of three glycosylation site mutants.
Two of the mutants, T24F and T40V, remove the aforementioned highly conserved but mutationally tolerant NGT and NAT glycosylation sites near the start of HA1.
The third mutation, S287A, removes a glycosylation site that is highly variable in nature but that our experiments suggest is under functional constraint.
The new Figure S6 shows that the T24F and T40V mutants grow to titers comparable to wildtype, supporting the conclusion that they are expendable at least in cell culture. 
However, the titer of the S287A is modestly lower than wildtype, supporting the conclusion that this glycan contributes to viral growth in cell culture.

One possible explanation is simply that cell culture does not fully capture the constraints in nature.
But a more interesting possibility is that conservation in nature is not expected to strictly parallel mutational tolerance.
The tendency for a site to change in nature depends on how much selection there is for the mutations.
For instance, Wei et al (\textit{Science Translational Medicine}, 2:24ra21, 2010) have pointed out that glycosylation motifs evolved rapidly in the globular head of HA from human H1 influenza in the period from 1918 to 2008 (seasonal H1N1), but changed very little in the same period in the swine H1 lineages descended from the 1918 virus.
The likely reason is differential strengths of selection: human influenza is under strong pressure from accumulated immune memory, whereas swine influenza is under less immune pressure because pigs are short lived and so are often naive when infected with influenza.
This fact highlights that even mutationally tolerant sites will only change rapidly if there is a benefit to doing so, while even somewhat constrained sites can change if there is sufficient pressure.
A strength of our experiments is that we have controlled the selection pressure (selection only for viral growth), and so can measure mutational tolerance in the absence of the confounding effects of immunity.

In addition to the data in Figure S6, we have added the following text:

\begin{quote}
\textsl{
The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.}
\end{quote}
}

@@

3) It is also surprising that the transmembrane domain seems to be highly plastic, a feature that is certainly not present in nature where it is highly conserved within H3 HAs.

\response{We have validated that one of these mutations in the transmembrane domain, C199(HA2)K, reaches reverse-genetics titers comparable to wildtype.
These data are in the new Figure S7.
In addition, various other studies have found that chimeric HAs with transmembrane domains from completely different proteins can still maintain HA cell surface expression and fusion activity (Melikyan et al 1999 Mol Biol Cell, Schroth-Diez et al 1998 JVI, Dong et al 1992 JVI, Roth et al 1986 JCB).

We posit that the disparity between the conservation of the transmembrane domain in nature and its apparent mutational tolerance in our experiments could arise at least in part from a lack of immune pressure for mutations in nature, as discussed in the response to the previous point.
}

### 2018-06-23 paper/manuscript.tex
responded to points 2-3 of reviewer 1

However, there are also some notable differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences.
Most surprisingly, the start codon at position does not show a particularly strong preference for methionine.
We validated that viruses carrying a mutation at this site from methionine to lysine do in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.
\comment{I AM HERE}
a notable exception occurs at the start codon at position -16, which does not show a strong preference for methionine (Figure~\ref{fig:logoplot}).
This codon is part of the signal peptide and is cleaved from the mature HA protein.
One possible reason that our experiments do not show a strong preference for methionine at this site could be alternative translation-initiation at a downstream or upstream start site, as has been described for other HAs~\cite{girard2011upstream}.

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
\caption{
\comment{Just show the first panel (A) here, the others can go in later figures}
{\bf Validation that the Perth/2009 HA is somewhat tolerant to mutation of the canonical start codon.}
We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation \comment{what was it}), a conserved cysteine was mutated to alanine (C52A, codon mutation \comment{what was it?}), and the same cysteine was synonymously mutated (C52C, codon mutation \comment{what was it}).
We selected these mutations because our deep mutational scanning results in Figure~\ref{fig:logoplot} surprisingly suggests that the start codon (position -16) is fairly tolerant of mutations, but that site 52 is highly intolerant of mutation to anything other than cysteine.
The synonymous C52C mutation is a negative control mutation that is not expected to have any effect.
We then used reverse genetics to generate viruses carrying the wildtype HA or each point mutant, with GFP packaged in the PB1 segment to enable easy titering by flow cytometry~\cite{bloom2010permissive,hooper2013mutant}.
Each variant was generated in duplicate.
The plots show the viral titer in the supernatant at 53 hours post-transfection.
As expected, the C52A virus is essentially non-viable, while the C52C mutation is roughly equivalent to wildtype.
the M(-16)K mutation is only moderately attenuated, validating that the canonical start codon is not completely essential for viral growth.
}
\label{suppfig:startcodon_validation}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
\caption{\label{XXX}
{\bf Viral titers of individual Perth/2009 H3 HA point mutants.}
Influenza GFP viruses carrying wildtype or single point mutants of the Perth/2009 HA were rescued by reverse genetics in duplicate. 
Viral transfection supernatant was harvested at 53 hours post-transfection and titered by flow cytometry in MDCK-SIAT1-TMPRSS2 cells.
Note that the wildtype, C52A and C52C mutant virus titers shown in each panel represent the same duplicate measurements.
The points mark each duplicate and the bar marks the mean.
(A) The start codon mutant M(-16)K reaches titers approximately one log less than those of wildtype. 
The C52A mutant fails to reach high titers, and the C52C synonymous variant grows to titers comparable to those of wildtype virus.
(B) The transmembrane domain mutant C199(HA2)K grows to higher titers than the wildtype.
(C) The T24F and T40V mutants, which remove glycosylation motifs, attain titers similar to or higher than wildtype, respectively. 
The S287A mutant, which also removes a glycosylation motif, does not grow as well as wildtype virus.
}
\end{suppfigure}

>>>>

However, there are also some notable differences between the amino-acid preferences measured in our experiments and amino-acid frequencies in natural H3 HA sequences.
Most surprisingly, the start codon does not show a particularly strong preference for methionine (Figure~\ref{fig:logoplot}).
We validated that a virus carrying a mutation at this site from methionine to lysine does in fact reach appreciable titers (Figure~\ref{suppfig:startcodon_validation}), perhaps because of alternative translation-initiation at a downstream or upstream start site as has been described for other HAs~\cite{girard2011upstream}.
Our measurements also suggest mutational tolerance at some other sites that are relatively conserved among natural HAs, such as the N-linked glycosylation motifs near the beginning of HA1 and the transmembrane domain (Figure~\ref{fig:logoplot}).
We validated that viruses with mutations to the glycosylation motifs at sites 22 or 38, or a site in the transmembrane domain, do in fact grow to high titers (Figures~\ref{suppfig:glycan_validation} and \ref{suppfig:TM_validation}, respectively).
The disparity between the relative conservation of these sites in nature and their mutational tolerance in our study could be because cell culture does not fully capture the constraints on HA function on nature, or could be because these sites are not under strong immune pressure and so mutations at them are not positively selected in nature.

@@

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
\caption{
\comment{Just show the first panel (A) here}
{\bf Validation that the Perth/2009 HA is somewhat tolerant to mutation of the canonical start codon.}
We created variants of the Perth/2009 HA in which the canonical start codon was mutated (amino-acid mutation M(-16)K, codon mutation \comment{what was it}), a conserved cysteine was mutated to alanine (C52A, codon mutation \comment{what was it?}), and the same cysteine was synonymously mutated (C52C, codon mutation \comment{what was it}).
We selected these mutations because our deep mutational scanning results in Figure~\ref{fig:logoplot} surprisingly suggests that the start codon (position -16) is fairly tolerant of mutations, but that site 52 is highly intolerant of mutation to anything other than cysteine.
The synonymous C52C mutation is a negative control mutation that is not expected to have any effect.
We then used reverse genetics to generate viruses carrying the wildtype HA or each point mutant, with GFP packaged in the PB1 segment to enable easy titering by flow cytometry~\cite{bloom2010permissive,hooper2013mutant}.
Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
As expected, the C52A virus is essentially non-viable, while the C52C mutation is roughly equivalent to wildtype.
the M(-16)K mutation is only moderately attenuated, validating that the canonical start codon is not completely essential for viral growth.
}
\label{suppfig:startcodon_validation}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
\caption{\label{suppfig:glycan_validation}
\comment{Just show the third panel (C) here, removing the cysteine mutants and adjusting the y-axis to a more appropriate for just these mutants.}
{\bf Validation of viral variants with mutations at N-linked glycosylation motifs.}
We created variants of the Perth/2009 HA with mutations that eliminated N-linked glycosylation motifs (Asn-Xaa-Ser/Thr) at asparagine residues 22, 38, and 285 (these are mutations T24F, T40V, and S287A, respectively).
We then used reverse genetics to generate viruses carrying the wildtype HA or each of these mutants.
Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
The viruses with mutations at the motifs at residues 22 and 38 reached titers at least as high as wildtype, whereas the virus with a mutation to the motif at residue 285 was modestly attenuated.
}
\end{suppfigure}

\begin{suppfigure}[H]
\centerline{\includegraphics[width=0.8\textwidth]{figs/supp_point_mut_validation/point_mut_validation.pdf}}
\caption{\label{suppfig:TM_validation}
\comment{Just show TM mutant here, similar to previous figure}
{\bf Validation of the mutational tolerance of a site in the transmembrane domain.}
We created a variant of the Perth/2009 HA with a transmembrane domain mutation, C199(HA2)K, at a site that our deep mutational scanning suggests should be highly mutationally tolerant (Figure~\ref{fig:logoplot}).
We then used reverse genetics to generate viruses carrying the wildtype HA or this mutants
Each variant was generated in duplicate, and the plots show the viral titer in the supernatant at 53 hours post-transfection.
The virus the mutation reached titers comparable to wildtype.
}
\end{suppfigure}
